A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-06 Simulation/Writeup/NoteConvergence.Rnw
pdf update

The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI methods. These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

>>>>

The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of this MI method. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

### 2019-12-06 Simulation/Writeup/NoteConvergence.tex
pdf update

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods. More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.

The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms. These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This paper only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

@@

\textbf{[Explain `simulation condition' (the number of iterations the algorithm has before imputing), ...?]}

@@

@@

High R hat would indicate .... low R hat values indicate ....

>>>>

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this research project is to investigate the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: `How to diagnose convergence of the multiple imputation algorithm implemented in \pkg{mice}?'.

The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of this MI method. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

@@

@@

The method of multiple imputation The properties of multiple imputation procedures have been studied

@@

High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
